### 2018-01-08 submission/manuscript.Rmd
incorporate RR genera updates

select_inc_4_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_stool.csv"))
select_inc_4_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_tissue.csv"))

adn_select_inc_4_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_stool.csv"))
adn_select_inc_4_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_tissue.csv"))

@@

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if the presence of specific genera resulted in a higher RR for both stool and tissue. To investigate this we analyzed the classically associated CRC genera, *Bacteroides*, *Escherichia*, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, *Porphyromonas*, and *Streptococcus* for an increase in RR. The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual. In fecal samples, the RR for carcinoma when having CRC-associated genera at a higher than overal study median relative abundance was greater than the RR associated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].

In stool the significant measures that increased RR of adenoma for CRC-associated genera were having a higher than overall study median relative abundance of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and being positive for three CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A]. Within tissue the significant measures for an increased RR of adenoma were being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though CRC-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These CRC-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.

The data presented herein supports the importance of CRC-associated genera for carcinoma but not necessarily adenoma formation. A proposed mechanism by which these CRC-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair. Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity [Figure 3C], are highly supportive of their importance. However, our observations show that when adding the community context to our Random Forest models along with the CRC-associated genera, the model AUC increases [Figure 4 & 5]. This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating CRC-associated microbes and carcinoma. Conversely, using the present data, it is less likely that CRC-associated microbes are correlated with adenoma development. The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself. The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of *Propionibacterium*, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination. The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC. Further, the data presented here suggests that non-resident CRC-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma. Conversely, no clear signal with these CRC-associated micobes could be detected for adenoma. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Overall, the microbiota associations with carcinoma are much stronger than the associations with adenoma.

>>>>

@@

***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue. To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3]. Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3]. Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.

For adenoma, there was almost no overlap between genera with a RR P-value of less than 0.05 and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had a P-value under 0.05 for either an increased or dereased RR are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both tissue and stool [Table S10 & S11]. Although the most significant increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.

The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation. A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair. Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance. However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5]. This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma. Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development. The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself. The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination. The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC. Further, the data presented here suggests that non-resident mouth-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma. Conversely, no clear signal with these mouth-associated micobes could be detected for adenoma. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Overall, the microbiota associations with carcinoma are much stronger than the associations with adenoma.

### 2018-01-08 submission/manuscript.tex
incorporate RR genera updates

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value \textgreater{} 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC-associated genera to build a prediction model resulted in
higher classification success, based on Area Under the Curve (AUC), for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma and one reason why this may be is that most studies were
only adequately powered for large effect sizes.

@@

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
metrics, we assessed if the presence of specific genera resulted in a
higher RR for both stool and tissue. To investigate this we analyzed the
classically associated CRC genera, \emph{Bacteroides},
\emph{Escherichia}, \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, \emph{Porphyromonas}, and
\emph{Streptococcus} for an increase in RR. The majority of
CRC-associated genera for both feces and tissue had a significantly
increased RR for carcinoma but not for adenoma {[}Figure 3{]}. The RR
effect size was greater for stool (RR range = 1.62 - 2.37) than for
tissue (RR range = 1.21 - 1.81). This decrease may be explained by the
fact that the tissue analysis included matched samples from the same
individual. In fecal samples, the RR for carcinoma when having
CRC-associated genera at a higher than overal study median relative
abundance was greater than the RR associated with evenness or Shannon
diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
continuously increased as individuals tested positive for more
CRC-associated genera {[}Figure 3B \& 3D{]}.

In stool the significant measures that increased RR of adenoma for
CRC-associated genera were having a higher than overall study median
relative abundance of \emph{Porphyromonas} (P-value = 0.023) and being
positive for three CRC-associated genera (P-value = 0.022) {[}Figure
3A{]}. Within tissue the significant measures for an increased RR of
adenoma were being positive for one CRC-associated genera (P-value =
0.032), being positive for two CRC-associated genera (P-value = 0.008),
and being positive for four CRC associated genera (P-value = 0.039)
{[}Figure 3C{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
CRC-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These CRC-associated genera are
clearly important to carcinoma classification but ourobservations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of CRC-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these CRC-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of CRC-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
CRC-associated genera positivity {[}Figure 3C{]}, are highly supportive
of their importance. However, our observations show that when adding the
community context to our Random Forest models along with the
CRC-associated genera, the model AUC increases {[}Figure 4 \& 5{]}. This
suggests the resident community cannot be ignored completely and needs
to be accounted for when investigating CRC-associated microbes and
carcinoma. Conversely, using the present data, it is less likely that
CRC-associated microbes are correlated with adenoma development. The
step-wise decrease in diversity suggests that the adenoma community is
not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the OTU models of
unmatched and matched {[}Figure S7E \& F{]} tissue samples were
concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only CRC-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the tissue results seem to be study specific with many top
10 taxa being present only in a single study. Additionally, the presence
of \emph{Propionibacterium}, within the top 10 most important variables
for the genera and OTU models, is worrying because it is commonly a
marker of contamination. The low bacterial biomass of tissue samples
coupled with potential contaminationcould ultimately explain why the
these results seem to be more sporadic than the stool results.

>>>>

\textbf{Results.} Fecal samples had a significant decrease from control
to adenoma to carcinoma for both Shannon diversity and evenness after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness resulted in small
increases in relative risk for adenoma (P-value = 0.032) and carcinoma
(P-value = 0.00034) while the reduction in Shannon diversity only
resulted in an increased relative risk for carcinoma (P-value = 0.0047).
Previously associated colorectal cancer genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
followed a similar pattern, with their presence significantly increasing
the relative risk for carcinoma (P-value \textless{} 0.05) but not
adenoma (P-value check tables 0.05) with the exception of
\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
only CRC-associated genera to build a prediction model resulted in
higher classification success, based on Area Under the Curve (AUC), for
both adenoma and carcinoma using fecal and tissue samples. The most
important OTUs for these models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Overall, there were less associations between the microbiota
and adenoma and one reason why this may be is that most studies were
only adequately powered for large effect sizes.

@@

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
RR of Lesion:}} Based on the small increase in RR using
\(\alpha\)-diversity metrics, we assessed if being higher than the
median for any specific genera resulted in an altered RR for adenoma or
carcinoma in stool and tissue. To investigate this we analyzed all
common genera across each study, in tissue or stool, and assessed
whether a relative abundance higher than the median results in an
increase or decrease in RR. For both stool and tissue none of the genera
in the top 5 most significantly increased RR were the same between
adenoma and carcinoma {[}Figure 3{]}. Mouth-associated genera were in
the top 5 genera associated with an increased RR of adenoma
(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and tissue.
Conversely, genera commonly associated with a normal gastrointestinal
tract were correlated with a decreased RR for both adenoma and carcinoma
for both stool and tissue {[}Figure 3{]}. Overall, there was little
direct overlap of the top 5 increased or decreased RR genera between
stool and tissue.

For adenoma, there was almost no overlap between genera with a RR
P-value of less than 0.05 and when they were similar the RR was in
opposite directions (e.g. \emph{Lactococcus}) {[}Table S8 \& S9{]}. Many
of the genera that had a P-value under 0.05 for either an increased or
dereased RR are also highly prevalent in contamination, specifically,
\emph{Novosphingobium}, \emph{Selemonas}, and \emph{Achromobacter}
{[}Figure 3 \& Table S8-S9{]}. For carcinoma, certain mouth-associated
genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}) had a high RR for both tissue and stool
{[}Table S10 \& S11{]}. Although the most significant increased RR
genera for tissue was \emph{Camplyobacter} while in stool it was
\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but our observations
suggest that accounting for the community in which these microbes exist
can increase the ability of models to make predictions.

The data presented herein supports the importance of mouth-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these mouth-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of mouth-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
mouth-associated genera positivity {[}Figure 3C{]}, are highly
supportive of their importance. However, our observations show that when
adding the community context to our Random Forest models along with the
mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
This suggests the resident community cannot be ignored completely and
needs to be accounted for when investigating mouth-associated microbes
and carcinoma. Conversely, using the present data, it is less likely
that mouth-associated microbes are correlated with adenoma development.
The step-wise decrease in diversity suggests that the adenoma community
is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the OTU models of
unmatched and matched {[}Figure S7E \& F{]} tissue samples were
concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
majority of the tissue results seem to be study specific with many top
10 taxa being present only in a single study. Additionally, the presence
of genera associated with contamination, within the top 10 most
important variables for the genera and OTU models, is worrying because
it is commonly a marker of contamination. The low bacterial biomass of
tissue samples coupled with potential contaminationcould ultimately
explain why the these results seem to be more sporadic than the stool
results.
